---
title: "Brandon EDA"
author: "Brandon Ritchie"
date: '2024-03-01'
output: html_document
---

```{r}
library(tidyverse)

# Read in Data
train_raw <- read_csv('../../Data/Train.csv') %>% 
  filter(age >= 18,# remove minors
         !is.na(income_above_limit)) %>% 
  mutate(income_above_limit = case_when(
    income_above_limit == 'Below limit' ~ 0,
    income_above_limit == 'Above limit' ~ 1
  )) 


```
First I will set a threshold of removing columns with more than 20% of records NA.

The only other columns that have values that represent NA's are the country of birth columns and the migration code columns. I think can remove country of birth columns in favor of the citizenship column.

```{r}
# Identify additional NA's
unique_values <- lapply(train_raw, unique)
#print(unique_values)

train_raw[train_raw == "?"] <- NA

# Clean up NA columns
na_cols <- train_raw %>% 
  summarise_all(~ sum(is.na(.)) / n())%>%
  pivot_longer(cols = everything(), names_to = 'Name', values_to = 'Value') %>% 
  filter(Value >= 0.20) %>% 
  pull(Name)

train1 <- train_raw %>% 
  select(-na_cols)




```


I performed the following cleaning on each column:

education: Education levels below high school are grouped because they are similar in output. The data has an ordinal order, but I decided to keep it categorical due to the likelihood of uneven spacing and interpretability. I also one-hot encoded the column for use in data modeling.
race: I added hispanic races and one-hot encoded
marital_status: Grouped present married couples (similar percentage) and left the rest. One-hot encoded the values.
working_week_per_year: There is a lot of noise in this data (a lot of 0), but a majority of the people that are classified in the target feature appeared to have a value above ~50. So, I encoded the column as a biary column of above 50 hours or not.
household_summary: Changed to a binary column of if they are the householder or not (largest seperation).
tax_status: One hot encoded.
citizenship: Changed into a binary column of whether they are a US citizen or not. Remove subsequent column about where they are from and where parents are from.




```{r}
# Education cleaning

# Needs to be grouped

new_order <- c(
  "Less than 1st grade",
  "1st 2nd 3rd or 4th grade",
  "5th or 6th grade",
  "7th and 8th grade",
  "9th grade",
  "10th grade",
  "11th grade",
  "12th grade no diploma",
  "High school graduate",
  "Some college but no degree",
  "Associates degree-occup /vocational",
  "Associates degree-academic program",
  "Bachelors degree(BA AB BS)",
  "Masters degree(MA MS MEng MEd MSW MBA)",
  "Doctorate degree(PhD EdD)",
  "Prof school degree (MD DDS DVM LLB JD)"
)

# Relevel the education_level column in df
train1$education <- factor(train1$education, levels = new_order)

train1 %>% 
  count(education)

# Columns Visualized
train1 %>% 
  ggplot(aes(age, color = as.factor(income_above_limit)))+
  geom_density(alpha = 0.2)+
  labs(title = 'Age Distribution by Target Classifier')

train1 %>% 
  group_by(education) %>% 
  summarise(perc_above_limit = sum(income_above_limit) / n()) %>% 
  ggplot(aes(education, perc_above_limit))+
  geom_col()+
  coord_flip()+
  labs(title = 'Percent Above Limit by Education')

train1 %>% 
  group_by(marital_status) %>% 
  summarise(perc_above_limit = sum(income_above_limit) / n()) %>% 
  ggplot(aes(marital_status, perc_above_limit))+
  geom_col()+
  coord_flip()+
  labs(title = 'Percent Above Limit by Marital Status')

train1 %>% 
  count(marital_status)

train1 %>% 
  group_by(race) %>% 
  summarise(perc_above_limit = sum(income_above_limit) / n()) %>% 
  ggplot(aes(race, perc_above_limit))+
  geom_col()+
  coord_flip()+
  labs(title = 'Percent Above Limit by Race')

train1 %>% 
  group_by(is_hispanic) %>% 
  summarise(perc_above_limit = sum(income_above_limit) / n()) %>% 
  ggplot(aes(is_hispanic, perc_above_limit))+
  geom_col()+
  coord_flip()+
  labs(title = 'Percent Above Limit by "Is Hispanic"')

train1 %>% 
  group_by(employment_commitment) %>% 
  summarise(perc_above_limit = sum(income_above_limit) / n()) %>% 
  ggplot(aes(employment_commitment, perc_above_limit))+
  geom_col()+
  coord_flip()+
  labs(title = 'Percent Above Limit by "Employment Commitment"')

train1 %>% 
  ggplot(aes(wage_per_hour))+
  geom_histogram()+
  labs(title = "Most Wage Per Hour values are 0")

train1 %>% 
  ggplot(aes(working_week_per_year, income_above_limit))+
  geom_jitter(alpha = 0.2)+
  labs(title = 'Working Week Per Year Breakdown')

train1 %>% 
  count(industry_code_main) %>% 
  mutate(perc = n / 150912) %>% 
  arrange(desc(perc))

train1 %>% 
  group_by(household_summary) %>% 
  summarise(perc_above_limit = sum(income_above_limit) / n()) %>% 
  ggplot(aes(household_summary, perc_above_limit))+
  geom_col()+
  coord_flip()+
  labs(title = 'Percent Above Limit by Household Summary')

# What does this mean?
train1 %>% 
  ggplot(aes(total_employed, income_above_limit))+
  geom_jitter(alpha = 0.2)

train1 %>% 
  group_by(tax_status) %>% 
  summarise(perc_above_limit = sum(income_above_limit) / n()) %>% 
  ggplot(aes(tax_status, perc_above_limit))+
  geom_col()+
  coord_flip()+
  labs(title = 'Percent Above Limit by Tax Status')

train1 %>% 
  ggplot(aes(gains))+
  geom_histogram()+
  labs(title = "Gains Dist")

train1 %>% 
  ggplot(aes(losses))+
  geom_histogram()+
  labs(title = "Losses Dist")

train1 %>% 
  ggplot(aes(stocks_status))+
  geom_histogram()+
  labs(title = "Stocks Status Dist")

train1 %>% 
  group_by(citizenship) %>% 
  summarise(perc_above_limit = sum(income_above_limit) / n()) %>% 
  ggplot(aes(citizenship, perc_above_limit))+
  geom_col()+
  coord_flip()+
  labs(title = 'Percent Above Limit by Citizenship')

# Renaming and grouping cleaning step
train2 <- train1 %>% 
  mutate(education = case_when(
    education %in% c("Less than 1st grade","1st 2nd 3rd or 4th grade",
"5th or 6th grade","7th and 8th grade","9th grade","10th grade",
"11th grade","12th grade no diploma") ~ "Less_than_HS",
education %in% c("Doctorate degree(PhD EdD)", "Prof school degree (MD DDS DVM LLB JD)") ~ "PhD_or_greater",
education %in% c('Associates degree-academic program', 'Associates degree-occup /vocational') ~ "Associates",
education == 'Bachelors degree(BA AB BS)' ~ 'Bachelors',
education == 'High school graduate' ~ "HS_Grad",
education == 'Masters degree(MA MS MEng MEd MSW MBA)' ~ 'Masters',
education == 'Some college but no degree' ~ 'Some_College',
is.na(education) ~ 'Unknown',
TRUE ~ education
  ),
race = case_when(
  race == 'Asian or Pacific Islander' ~ 'Asian',
  race == 'Amer Indian Aleut or Eskimo' ~ 'Native_American',
  !(is_hispanic %in% c(NA, 'Do not know', 'All other')) ~ 'Hispanic', 
  TRUE ~ race
),
marital_status = case_when(
  marital_status %in% c('Married-A F spouse present', 'Married-civilian spouse present') ~ 'Married_Spouse_Present',
  marital_status == 'Married-spouse absent' ~ 'Married_Spouse_Absent',
  marital_status == 'Never married' ~ 'Unmarried',
  TRUE ~ marital_status
),
week_working_hours_above_50 = case_when(
  working_week_per_year >= 50 ~ 1,
  TRUE ~ 0
),
head_of_house = case_when(
  household_summary == 'Householder' ~ 1,
  TRUE ~ 0
),
tax_status = case_when(
  tax_status %in% c('Joint both 65+', 'Joint one under 65 & one 65+') ~ 'Joint_>=65',
  tax_status == 'Joint both under 65' ~ 'Joint_<65',
  tax_status == 'Head of household' ~ 'HoH',
  TRUE ~ tax_status
),
us_citizen = case_when(
  citizenship == 'Foreign born- Not a citizen of U S' ~ 0,
  TRUE ~ 1
),
employment_status = case_when(
  employment_commitment %in% c("Unemployed full-time", "Unemployed part- time", "Not in labor force") ~ 'Unemployed',
  TRUE ~ "Employed"
))

# One-hot encoding
cleaned_train <- cbind(train2, 
      model.matrix(~ education - 1, data = train2), 
      model.matrix(~ gender - 1, data = train2), 
      model.matrix(~ race - 1, data = train2),
      model.matrix(~ marital_status - 1, data = train2),
      model.matrix(~ tax_status - 1, data = train2)) %>% 
  select(-c(ID, 
            education, 
            gender, 
            is_hispanic, 
            race, 
            marital_status, 
            employment_commitment, # Review this
            employment_stat, # Removing, because I am not sure what this means
            wage_per_hour, # Removing because not sure what mean (scale is off and most values are 0)
            working_week_per_year,
            occupation_code, # Not sure what this represents
            industry_code_main, # 30% of data is unknown
            household_summary,
            household_stat,
            total_employed, # Not sure what this means
            tax_status,
            industry_code, 
            mig_year,
            gains, # Not enough data
            losses, # Not enough data
            stocks_status, # Not enough data
            importance_of_record, # Don't know what means
            country_of_birth_own,
            country_of_birth_father,
            country_of_birth_mother,
            vet_benefit,
            citizenship
            )) 
  
employed_train <- cleaned_train %>% 
  filter(employment_status == 'Employed')

unemployed_train <- cleaned_train %>% 
  filter(employment_status == 'Unemployed')

```

Cleaned dataset:

```{r}
employed_train %>% 
  head(10)

unemployed_train %>% 
  head(10)
```




